Drought frequency predicts life history strategies in \emph{Heliophila}

### 2018-12-12 manuscript/manuscript.tex
submission

\abstract{
Explaining variation in life history strategies is a long-standing goal
of evolutionary biology. For plants, annual and perennial life histories
are thought to reflect adaptation to environments that differ in the
frequency of stress events such as drought. Here we test this hypothesis
in \textit{Heliophila} (Brassicaceae), a diverse genus of flowering
plants native to Africa, by integrating 34 years of satellite-based
drought measurements with 2192 herbaria occurrence records. Consistent
with predictions from classic life history theory, we find that
perennial \textit{Heliophila} species occur in environments where
droughts are significantly less frequent compared to annuals. These
associations are predictive while controlling for phylogeny, lending
support to the hypothesis that drought related natural selection has
influenced the distributions of these strategies. Additionally, the
collection dates of annual and perennial species indicate that annuals
escape drought prone seasons during the seed phase of their life cycle.
Together, these finding provide empirical support for classic hypotheses
about the drivers of life history strategy in plants - that perennials
out compete annuals in environments with less frequent drought and that
annuals are adapted to environments with more frequent drought by
escaping drought prone seasons as seeds. 
}

@@

Understanding the causes and consequences of life history variation is a
longstanding goal of ecology and evolutionary biology (Cole, 1954). In
plants, life histories are especially diverse, with herbaceous species
that complete their life cycle in a number of weeks to trees that live
for thousands of years (Brown, 1996). Along this continuum an important
division exists, distinguishing annuals which complete their seed to
seed life cycle within a single calendar year from perennials which can
persist over multiple years. Annual plants flower once, set seed,
senesce, and then die, spending at least some portion of the year as a
seed, where they are relatively protected from environmental stress. In
contrast, perennial plants can continue vegetative growth after
reproduction and must survive conditions experienced during all seasons.
These represent fundamentally different life history strategies, but the
ecological factors that explain their evolution and distributions remain
empirically uresolved (Friedman \& Rubin, 2015).\\
Classical theory predict shorter life spans in environments where adult
mortality is high (Charnov \& Schaffer, 1973; Stearns, 1992; Franco \&
Silvertown, 1996). In plants, this has been extended to the hypothesis
that annuality is adaptive when it allows plants to escape drought
(Schaffer \& Gadgil, 1975). Lack of water is perhaps the greatest threat
to survival during vegetative or reproductive growth and annuals can
remain dormant (and protected as a seed) during drought. Thus,
environments with greater seasonal drought frequency may select for
annual life histories that complete reproduction prior to drought prone
seasons. Conversely, environments with less frequent drought may select
for perennial species, which may benefit from multiple bouts of
reproduction and competitive advantage by preventing recruitment of
annual species (Corbin \& D'Antonio, 2004). These predictions have been
supported by the observation of annuals in arid environments in
\emph{Oryza perennis} (Morishima \emph{et al.}, 1984) and
\emph{Oenothera} (Evans \emph{et al.}, 2005). Additionally, annual and
perennial species of \emph{Nemesia} were qualitatively associated with
winter rather and summer rainfall environments respectively (Datson
\emph{et al.}, 2008) and annual species of \emph{Scorzoneroides} were
associated with environments classified as unpredictable (Cruz-Mazo
\emph{et al.}, 2009). However, whether the history frequency of drought
events indeed predicts the distributions annual or perennial life
history strategies has yet to be tested.\\
Here we combine a long-term global dataset of satellite detected drought
events with metadata from natural history collections to test these
classic hypotheses within the African endemic mustard genus,
\emph{Heliophila} L. (Brassicaceae). If annuality is an adaptive
strategy allowing plants to escape drought prone seasons, then drought
frequency should predict the distribution of life history strategies
across landscapes, and annual species should be more commonly associated
with drought prone regions than perennial species. Furthermore, if
annual species have adapted to escape drought prone seasons,
observations of growing annual species (i.e.~occurring in forms other
than seed) should be rare during drought prone seasons. Phylogenetic
relatedness can influence tests of associations between species' traits
and their environments (Felsenstein, 1985; Barrett \emph{et al.}, 1996),
and therefore we assessed the relationship between life history
distribution and drought frequency in a phylogenetic context.

@@

Botanists have collected and maintained over 350 million botanical
specimens worldwide over the past 300 years. Herbarium specimens and
their associated metadata have been used since the 1960s to study
species' geographical distributions (reviewed by Willis \emph{et al.}
(2017) and Lang \emph{et al.} (2018)). And as they become digitized
(Soltis, 2017), these collections have been used to study relationships
between trait distributions, geography, and climate (Davis \emph{et
al.}, 2015; Stropp \emph{et al.}, 2016; Wolf \emph{et al.}, 2016;
Václavı'k \emph{et al.}, 2017). To characterize the distributions of
annual and perennial \emph{Heliophila} species, all records for the
genus \emph{Heliophila} were downloaded from the Global Biodiversity
Information Facility (gbif.org) on July 21, 2018 (GBIF, 2018).

@@

\begin{itemize}
\tightlist
\item
  records for species with reported life history\\
\item
  records with geospatial data\\
\item
  records without known geospatial coordinate issues (i.e., coordinates
  reported are those of herbarium)\\
\item
  records from collection sites classified as land pixels\\
\item
  records from Africa (to exclude locations of cultivation)
\item
  records without duplicates (i.e., identical species, location,
  collection date)
\end{itemize}

@@

To evaluate the hypothesis that annual and perennial life history
strategies reflect adaptations to alternative drought regimes, we tested
the corresponding prediction that the observed distributions of annual
and perennial \emph{Heliophila} species would be significantly
associated with historic drought frequency. First, we compared the
frequency of drought during the winter, spring, summer, and fall between
raw occurrence records of annual and perennial species by t-tests. To
account for variation in the number of occurrence records per species,
we next calculated the mean drought frequency during the winter, spring,
summer and fall for each species. Because shared evolutionary history of
closely related species can lead to spurious associations between traits
and environments (Felsenstein, 1985), we tested for a relationship
between life history strategy and drought frequency while controlling
for phylogeny using phylogenetic logistic regression (Ives \& Garland,
2010).

@@

Out of 8670 \emph{Heliophila} GBIF records, 6634 were for species with
reported life history (Mummenhoff \emph{et al.}, 2005), 2856 had
geospatial data, 2833 did not have geospatial issues, 2684 were located
on pixels classified as land having drought measurements, 2543 were
located in Africa, 2192 were not duplicated. Thus, after all filtering
steps, 2192 records for 42 species (Figure \ref{fig:phylogeny}, Table
\ref{tab:speciesmeanstable}) passed for further analyses. The number of
samples varied between species, with a mean of 52.19 samples per
species. Heliophila rigidiuscula had the most records, 201, and
Heliophila cornellsbergia the fewest, 2 (Table
\ref{tab:speciesmeanstable}).

@@

Theory predicts that annuality should be adaptive in places where
stresses such as drought are more common. Conversely, perenniality
should be adaptive in places where such stresses are less frequent. We
found that the frequency of drought was significantly higher at the
locations of occurrence records for annual species. In terms of raw
observation records (Figure \ref{fig:mapsboxplots}g), the frequency of
drought was significantly higher at the location of annuals during the
winter (t = 10.65, p = 0.00), spring (t = 10.73, p = 0.00), summer (t =
12.67, p = 0.00), and fall (t = 15.26, p = 0.00). Because raw occurrence
records do not account for variation in the number of records per
species (Table \ref{tab:speciesmeanstable}) or species relatedness
(Figure \ref{fig:phylogeny}a), we also tested whether mean drought
frequency values of each species were significantly different between
annuals and perennials using phylogenetic logistic regression. We found
that the mean drought frequencies were significantly higher
(\(\alpha = 0.05\)) in annual species during the spring, summer, and
fall (Table \ref{tab:modelstable}, Figure \ref{fig:lineplots}a). These
findings indicate that common acestry alone does not explain differences
the drought frequencies experienced in the environments of annual and
perennial \emph{Heliophila}.

\begin{table}[tbp]
\begin{center}
\begin{threeparttable}
\caption{\label{tab:modelstable}Phylogenetic logistic regressions between life history, and the mean drought frequency observed at herbaria collection sites of Heliophila species the winter, spring, summer, and fall.}
\begin{tabular}{lll}
\toprule
Predictor & \multicolumn{1}{c}{Estimate} & \multicolumn{1}{c}{P}\\
\midrule
Intercept & 0.7231 & 0.6636\\
Winter drought freq. & -1.5452 & 0.7274\\ \midrule
Intercept & 5.0107 & 0.0534\\
Spring drought freq. & -12.9014 & 0.0464\\ \midrule
Intercept & 7.7093 & 0.0054\\
Summer drought freq. & -19.9056 & 0.0042\\ \midrule
Intercept & 7.0162 & 0.0082\\
Fall drought freq. & -20.8174 & 0.0067\\ \midrule
\bottomrule
\addlinespace
\end{tabular}
\begin{tablenotes}[para]
\normalsize{\textit{Note.} Annual species were scored as 0 and perennial species as 1.}
\end{tablenotes}
\end{threeparttable}
\end{center}
\end{table}

@@

The preceding results indicated that annual species are found in
environments where droughts are significantly more frequent, especially
in the summer and fall. Classic life history theory hypothesizes that
annuality reflects adaptation to such environments because it allows
species to escape stressful conditions. If this is the case, we would
expect that annuals spend the drought prone seasons of summer and fall
as seeds. To test this hypothesis, we compared the dates of occurrence
records between annual and perennial \emph{Heliophila} species. The
distributions reveal a considerable difference in the timing of
observation of these two life histories. In comparison to perennials,
which appear to be collected throughout the year, annuals are almost
exclusively observed during the winter and spring (Figure
\ref{fig:lineplots}b). The differences between the distribution of
collection dates were significant by all tests (ks.test D = 0.25, p = 0;
bartlett.test K2 = 503.18, p = 0.00) This is consistent with a model of
life history in which annual species flower in the spring, set seed,
senesce, and die before the summer. Thus, these annual species are
likely to remain dormant during the summer and fall, when drought is the
strongest predictor of the distributions of annual and perennial life
histories (Figure \ref{fig:lineplots}a).

@@

These findings support classical theoretical predictions about the
adaptive value of annual and perennial life history strategies. Taken
together, they suggest that in \emph{Heliophila}, annual species are
adapted to environments with increased summer droughts by avoiding these
seasons in a dormant seed phase of their life cycle. They also suggest
that perenniality is adaptive in environemnts where droughts are less
frequent. While most previous work has focused on describing the
evolutionary origins of annuality (Barrett \emph{et al.}, 1996; Conti
\emph{et al.}, 1999; Andreasen \& Baldwin, 2001; Verboom \emph{et al.},
2004; Friedman \& Rubin, 2015) there are at least a few other cases
where perenniality appears to have arisen from an annual ancestor (Bena
\emph{et al.}, 1998; Tank \& Olmstead, 2008). And while early theory
predicted selection for annuality when adult morality is high (Stearns,
1992), we also find evidence that the transition to perenniality could
be explained by the historical frequency of drought, a likely cause of
mortality in plants. The phylogeny reveals several transitions from
annual to perennial life history (Figure \ref{fig:phylogeny}a) and that
the distributions of perennial \emph{Heliophila} extend into regions
where drought frequency is low (Figure \ref{fig:mapsboxplots}b, Figure
\ref{fig:speciesmaps}). Perennials may be able to out complete annual
relatives in environments where the infrequency of drought favors
strategies that allow plants to benefit from growth over many seasons.
It may also indicate that annuals rely on drought as a source of
disturbance for seedling recruitment when competing with perennials
(Corbin \& D'Antonio, 2004). Indeed, no annual species were observed in
the low drought regions of eastern South Africa (Figure
\ref{fig:mapsboxplots}, Figure \ref{fig:speciesmaps}).

@@

We thank Jesse Lasky and members of the Sloan lab for generous feedback
that improved the quality of this work. This research was supported by
NSF Award 1701918 and USDA-NIFA Award 2014-38420-21801 to JGM.

@@

\leavevmode\hypertarget{ref-morishima1984differentiation}{}%
\textbf{\textnormal{Morishima H}, \textnormal{Sano Y}, \textnormal{Oka
H}}. \textbf{1984}. Differentiation of perennial and annual types due to
habitat conditions in the wild riceOryza perennis. \emph{Plant
Systematics and Evolution} \textbf{144}: 119--135.

@@

>>>>

\abstract{
Explaining variation in life history strategies is a long-standing goal
of evolutionary biology. For plants, annual and perennial life histories
are thought to reflect adaptation to environments that differ in the
frequency of stress events such as drought. Here we test this hypothesis
in \textit{Heliophila} (Brassicaceae), a diverse genus of flowering
plants native to Africa, by integrating 34 years of satellite-based
drought measurements with 2192 herbaria occurrence records. Consistent
with predictions from classic life history theory, we find that
perennial \textit{Heliophila} species occur in environments where
droughts are significantly less frequent compared to annuals. These
associations are predictive while controlling for phylogeny, lending
support to the hypothesis that drought related natural selection has
influenced the distributions of these strategies. Additionally, the
collection dates of annual and perennial species indicate that annuals
escape drought prone seasons during the seed phase of their life cycle.
Together, these findings provide empirical support for classic
hypotheses about the drivers of life history strategy in plants - that
perennials out compete annuals in environments with less frequent
drought and that annuals are adapted to environments with more frequent
drought by escaping drought prone seasons as seeds.

}

@@

Understanding the causes and consequences of life history variation is a
longstanding goal of ecology and evolutionary biology (Cole, 1954). In
plants, life histories are especially diverse, with herbaceous species
that complete their life cycle in a number of weeks to trees that live
for thousands of years (Brown, 1996). Along this continuum in
angiosperms an important division exists, distinguishing annuals which
complete their seed to seed life cycle within a single calendar year
from perennials which can persist over multiple years. Annual plants
flower once, set seed, senesce, and then die, spending at least some
portion of the year as a seed, where they are relatively protected from
environmental stress. In contrast, perennial plants can continue
vegetative growth after reproduction and must survive conditions
experienced during all seasons. These represent fundamentally different
life history strategies, but the ecological factors that explain their
evolution and distributions remain empirically uresolved (Friedman \&
Rubin, 2015).\\
Classical theory predicts shorter life spans in environments where adult
mortality is high (Charnov \& Schaffer, 1973; Stearns, 1992; Franco \&
Silvertown, 1996). In plants, this has been extended to the hypothesis
that annuality is adaptive when it allows plants to escape drought
(Schaffer \& Gadgil, 1975). Lack of water is perhaps the greatest threat
to survival during vegetative or reproductive growth and annuals can
remain dormant (and protected as a seed) during drought. Thus,
environments with greater seasonal drought frequency may select for
annual life histories that complete reproduction prior to drought prone
seasons. Conversely, environments with less frequent drought may select
for perennial species, which benefit from multiple bouts of reproduction
and competitive advantage by preventing recruitment of annual species
(Corbin \& D'Antonio, 2004). These predictions have been supported by
the observation of annuals in arid environments in \emph{Oryza perennis}
(Morishima \emph{et al.}, 1984) and \emph{Oenothera} (Evans \emph{et
al.}, 2005). Additionally, annual and perennial species of
\emph{Nemesia} were qualitatively associated with winter rather and
summer rainfall environments respectively (Datson \emph{et al.}, 2008)
and annual species of \emph{Scorzoneroides} were associated with
environments classified as unpredictable (Cruz-Mazo \emph{et al.},
2009). However, whether the history frequency of drought events indeed
predicts the distributions annual or perennial life history strategies
has yet to be tested.\\
Here we combine a long-term global dataset of satellite detected drought
events with metadata from natural history collections to test these
classic hypotheses within the African endemic mustard genus,
\emph{Heliophila} L. (Brassicaceae). If annuality is an adaptive
strategy allowing plants to escape drought prone seasons, then drought
frequency should predict the distribution of life history strategies
across landscapes, and annual species should be more commonly associated
with drought prone regions than perennial species. Furthermore, if
annual species have adapted to escape drought prone seasons,
observations of growing annual species (i.e.~occurring in forms other
than seed) should be rare during drought prone seasons. Phylogenetic
relatedness can influence tests of associations between species' traits
and their environments (Felsenstein, 1985; Barrett \emph{et al.}, 1996),
and therefore we assessed the relationship between life history
distribution and drought frequency in a phylogenetic context.

@@

Botanists have collected and maintained over 350 million botanical
specimens worldwide over the past 300 years (Thiers, 2016). Herbarium
specimens and their associated metadata have been used since the 1960s
to study species' geographical distributions (reviewed by Willis
\emph{et al.} (2017) and Lang \emph{et al.} (2018)). And as they become
digitized (Soltis, 2017), these collections have been used to study
relationships between trait distributions, geography, and climate (Davis
\emph{et al.}, 2015; Stropp \emph{et al.}, 2016; Wolf \emph{et al.},
2016; Václavı'k \emph{et al.}, 2017). To characterize the distributions
of annual and perennial \emph{Heliophila} species, all records for the
genus \emph{Heliophila} were downloaded from the Global Biodiversity
Information Facility (gbif.org) on July 21, 2018 (GBIF, 2018).

@@

\begin{itemize}
\tightlist
\item
  records for species with reported life history\\
\item
  records with geospatial data\\
\item
  records without known geospatial coordinate issues (i.e., coordinates
  reported are those of herbarium)\\
\item
  records from collection sites classified as land pixels in the VHI
  dataset\\
\item
  records from Africa (to exclude locations of cultivation)
\item
  records without duplicates (i.e., identical species, location,
  collection date)
\end{itemize}

@@

To evaluate the hypothesis that annual and perennial life history
strategies reflect adaptations to alternative drought regimes, we tested
the corresponding prediction that the observed distributions of annual
and perennial \emph{Heliophila} species would be significantly
associated with historic drought frequency. First, we compared the
frequency of drought during the winter, spring, summer, and fall between
total occurrence records of annual and perennial species by t-tests. To
account for variation in the number of occurrence records per species,
we next calculated the mean drought frequency during the winter, spring,
summer and fall for each species. Because shared evolutionary history of
closely related species can lead to spurious associations between traits
and environments (Felsenstein, 1985), we tested for a relationship
between life history strategy and drought frequency while controlling
for phylogeny using phylogenetic logistic regression (Ives \& Garland,
2010).

@@

Out of 8670 \emph{Heliophila} GBIF records, 6634 were for species with
reported life history (Mummenhoff \emph{et al.}, 2005), 2856 had
geospatial data, 2833 did not have geospatial issues, 2684 were located
on pixels classified as land having drought measurements, 2543 were
located in Africa, 2192 were not duplicated. Thus, after all filtering
steps, 2192 records for 42 species (Figure \ref{fig:phylogeny}, Table
\ref{tab:speciesmeanstable}) passed for further analyses. The number of
samples varied between species, with a mean of 52.19 samples per
species. \emph{H. rigidiuscula} had the most records, 201, and \emph{H.
cornellsbergia} the fewest, 2 (Table \ref{tab:speciesmeanstable}).

@@

Theory predicts that annuality should be adaptive in places where
stresses such as drought are more common. Conversely, perenniality
should be adaptive in places where such stresses are less frequent. We
found that the frequency of drought was significantly higher at the
locations of occurrence records for annual species. When comparing
across all occurrence records (all records rather than species means,
Figure \ref{fig:mapsboxplots}g), the frequency of drought was
significantly higher at the location of annuals during the winter (t =
10.65, p = 0.00), spring (t = 10.73, p = 0.00), summer (t = 12.67, p =
0.00), and fall (t = 15.26, p = 0.00). Because a comparison across all
occurrence records does not account for variation in the number of
records per species (Table \ref{tab:speciesmeanstable}) or species
relatedness (Figure \ref{fig:phylogeny}a), we also tested whether mean
drought frequency values of each species were significantly different
between annuals and perennials using phylogenetic logistic regression.
We found that the mean drought frequencies were significantly higher
(\(\alpha = 0.05\)) in annual species during the spring, summer, and
fall (Table \ref{tab:modelstable}, Figure \ref{fig:lineplots}a). These
findings indicate that common acestry alone does not explain differences
the drought frequencies experienced between the environments of annual
and perennial \emph{Heliophila}.

\begin{table}[tbp]
\begin{center}
\begin{threeparttable}
\caption{\label{tab:modelstable}Phylogenetic logistic regressions between life history, and the mean drought frequency observed at specimen sites of Heliophila species the winter, spring, summer, and fall.}
\begin{tabular}{lll}
\toprule
Predictor & \multicolumn{1}{c}{Estimate} & \multicolumn{1}{c}{P}\\
\midrule
Intercept & 0.7231 & 0.6636\\
Winter drought freq. & -1.5452 & 0.7274\\ \midrule
Intercept & 5.0107 & 0.0534\\
Spring drought freq. & -12.9014 & 0.0464\\ \midrule
Intercept & 7.7093 & 0.0054\\
Summer drought freq. & -19.9056 & 0.0042\\ \midrule
Intercept & 7.0162 & 0.0082\\
Fall drought freq. & -20.8174 & 0.0067\\ \midrule
\bottomrule
\addlinespace
\end{tabular}
\begin{tablenotes}[para]
\normalsize{\textit{Note.} Annual species were scored as 0 and perennial species as 1.}
\end{tablenotes}
\end{threeparttable}
\end{center}
\end{table}

@@

The preceding results indicate that annual species are found in
environments where droughts are significantly more frequent, especially
in the summer and fall. Classic life history theory hypothesizes that
annuality reflects adaptation to such environments because it allows
species to escape stressful conditions. If this is the case, we would
expect that annuals spend the drought prone seasons of summer and fall
as seeds. To test this hypothesis, we compared the dates of occurrence
records between annual and perennial \emph{Heliophila} species. The
distributions reveal a considerable difference in the timing of
observation of these two life histories. In comparison to perennials,
which appear to be collected throughout the year, annuals are almost
exclusively observed during the winter and spring (Figure
\ref{fig:lineplots}b). The differences between the distribution of
collection dates were significant by all tests (ks.test D = 0.25, p = 0;
bartlett.test K2 = 503.18, p = 0.00) This is consistent with a model of
life history in which annual species flower in the spring, set seed,
senesce, and die before the summer. Thus, these annual species are
likely to remain dormant during the summer and fall, when drought is the
strongest predictor of the distributions of annual and perennial life
histories (Figure \ref{fig:lineplots}a).

@@

These findings support classical theoretical predictions about the
adaptive value of annual and perennial life history strategies. Taken
together, they suggest that in \emph{Heliophila}, annual species are
adapted to environments with increased summer droughts by avoiding these
seasons in a dormant seed phase of their life cycle. They also suggest
that perenniality is adaptive in environemnts where droughts are less
frequent. While most previous work has focused on describing the
evolutionary origins of annuality (Barrett \emph{et al.}, 1996; Conti
\emph{et al.}, 1999; Andreasen \& Baldwin, 2001; Verboom \emph{et al.},
2004; Friedman \& Rubin, 2015) there are at least a few other cases
where perenniality appears to have arisen from an annual ancestor (Bena
\emph{et al.}, 1998; Tank \& Olmstead, 2008). And while early theory
predicted selection for annuality when adult morality is high (Stearns,
1992), we also find evidence that perenniality could be explained by
reduced frequency of drought. The phylogeny reveals several transitions
from annual to perennial life history (Figure \ref{fig:phylogeny}a) and
the distributions of perennial \emph{Heliophila} extend into regions
where drought frequency is low (Figure \ref{fig:mapsboxplots}b, Figure
\ref{fig:speciesmaps}). Perennials may be able to out complete annual
relatives in environments where the infrequency of drought favors
strategies that allow plants to benefit from growth over many seasons.
This also suggests that annuals rely on drought as a source of
disturbance for seedling recruitment when competing with perennials
(Corbin \& D'Antonio, 2004). Indeed, no annual species were observed in
the low drought regions of eastern South Africa (Figure
\ref{fig:mapsboxplots}, Figure \ref{fig:speciesmaps}).

@@

We thank members of the Sloan lab for generous feedback that improved
the quality of this work. This research was supported by NSF Award
1701918 and USDA-NIFA Award 2014-38420-21801 to JGM.

@@

\leavevmode\hypertarget{ref-morishima1984differentiation}{}%
\textbf{\textnormal{Morishima H}, \textnormal{Sano Y}, \textnormal{Oka
H}}. \textbf{1984}. Differentiation of perennial and annual types due to
habitat conditions in the wild rice oryza perennis. \emph{Plant
Systematics and Evolution} \textbf{144}: 119--135.

@@

\leavevmode\hypertarget{ref-thiers2016index}{}%
\textbf{\textnormal{Thiers B}}. \textbf{2016}. Index herbariorum: A
global directory of public herbaria and associated staff. New york
botanical garden's virtual herbarium. \emph{http://sweetgum. nybg.
org/ih}.
